Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-31 Multiresolution_framework/ltx/manuscript.tex
minor edits

The single layer homogeneous neural field model in the most general form is described by the equation
\begin{equation}
	\label{FullDoubleIntModel} \underbrace{v\left(\mathbf{r},t\right)}_{\text{field}} =
	\int_{-\infty}^t 
	\underbrace{h\left(t - t'\right)}_{\text{synapse}} \int_\Omega
	\theta\underbrace{w\left(\mathbf{r},\mathbf{r}'\right)}_{\text{connectivity}} 
	\underbrace{f\left( v\left( \mathbf{r}',t' \right)\right)}_{\text{firing}}
	\, \mathrm{d}\mathbf{r}'\mathrm{d}t',
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ within the spatial domain $\Omega$ is denoted by $v\left(\mathbf r,t\right)$. The function $h(t)$ describes the post-synaptic response kernel, $w(\mathbf{r})$ describes the connectivity kernel and $f(v(\mathbf{r},t))$ describes the relationship between the mean membrane potential and the mean firing rate. The parameter $\theta$ can be thought of the synaptic efficiency or coupling strength, where $\theta>0$ and $\theta<0$ describes excitation and inhibition, respectively. There are various forms of $h(t)$ that have been proposed in the literature, with different levels of complexity. These are shown in Figure~\ref{fig:Modelcomponents}. Similarly, both the connectivity kernel and the activity function (firing rate) can take various forms. These are also shown in Figure~\ref{fig:Modelcomponents}. The form of $w(\mathbf{r})$ typically describes an Laplacian (exponential) or Gaussian decay in connectivity strength from any point in the field. By assuming the form of $h(t)$ is the same across layers (scales) the Amari style neural field model is formed, incorporating a multilayer structure into the connectivity kernel. In this way, the connectivity kernel typically takes the form of a Mexican hat (Gaussian shape within layer connectivity) or Wizard hat (Laplacian shape within layer) as depicted in Figure~\ref{fig:Modelcomponents}.

@@

The $m$-th  order cardinal B-spline function is defined by the recurrence relation \citep{Chui1992} 
\begin{equation}
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)\,\mathrm{d}r',
\label{SplineConvolutionIntegral}
\end{equation}
where $m>1$, $\ast$ denotes convolution, and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$, also known as indicator function, i.e.
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Following this, a B-spline of any order, $N_m(r)$, can be computed using the recursive expression defined by \citep{DeBoor2001}
\begin{equation}\label{eq:MRA-DoBoorFormula}
 N_{m}\left(r\right)=\frac{r}{m-1}N_{m-1}\left(r\right)+\frac{m-r}{m-1}N_{m-1}\left(r-1\right) \quad m>1.
 \end{equation}
The most commonly used form of splines is the cubic spline $\left(m=4\right)$, comprised of third degree polynomials added together at the joining points, with an explicit expression given by
\begin{align}
3!N_{4}\left(r\right)=
\begin{cases}
r^3 & \text{if $ 0\le r\le1$}, \\
4-12r+12r^2-3r^3 & \text{if $1\le r\le2$}, \\
-44+60r-24r^2+3r^3 & \text{if $2\le r\le3$}, \\
64-48r+12r^2-r^3 & \text{if $3\le r\le4$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{align}
One important feature of the B-spline functions is the \emph{partition of unity} property, meaning that the unity can be expressed as a linear sum of B-splines, i.e
\begin{equation}
	\sum_{l}N_m(r+l)\equiv1
	\end{equation}
	This is important as the underlying field can be reconstructed in saturation without introducing numerical errors which is not the case for example in the case of Gaussian basis functions.

@@

By exploiting \eqref{eq:BsplineConvolution} and \eqref{eq:BsplineInnerProduct}, the integrals in \eqref{eq:Lambdax}, \eqref{eq:Lambdatheta} and \eqref{eq:CovMatrix} can be computed analytically. In constructing $\boldsymbol\Lambda_{x}$, it is important to note that B-spline scaling and wavelet functions possess the following orthogonality properties \citep{Unser1993}: 
\begin{equation}
  \left\langle \varphi_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\neq j_2
 \label{eq:MRA-PsiPsiOrthogonality} 
 \end{equation}
 \begin{equation}
  \left\langle N_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\leq j_2.
 \label{eq:MRA-PhiPsiOrthogonality}
 \end{equation}
  \parham{Note from \eqref{eq:MRA-PsiPsiOrthogonality} Bspline wavelets are semiorthogonal as they are not orthogonal with respect to translation on a given scale.}

@@

\subsection{Frequency Analysis}
To determine the level of approximation in the model, detailed in the previous section, that is the number of basis functions required to reconstruct the field, $v_t(r)$, the frequency response of the B-spline function needs to be computed by applying the Fourier transform to \eqref{eq:N1convolutions}. To proceed the Fourier transform of $N_1(r)$ is first calculated, 
\begin{align}\label{eq:MRA-N1Fouriertransform}
\mathcal F(N_1(r))&=\int_{-\infty}^{+\infty}N_1(r)\mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\int_{0}^{1} \mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}.
\end{align}
The convolution theorem states that the Fourier transform of the convolution of functions is equal to the product of their Fourier transforms. Therefore, taking Fourier transform of \eqref{eq:N1convolutions} and substituting \eqref{eq:MRA-N1Fouriertransform} for $\mathcal F(N_1(r)) $ yields
\begin{align}\label{eq:MRA-NmFouriertransform}
\mathcal F(N_m(r))=\left(\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}\right)^m.
\end{align}
\dean{Need to say how the FT is used somewhere here before proceeding.} \parham{The Fourier transform of the $m$-th order Bspline allows the calculation of the spatial frequency response of the wavelets using \eqref{eq:MRA-TwoScalepair2} from which the level of approximation, $j$, can be determined. This way we choose wavelets upto level $j$ which fully describe the spatial frequency contents of the underlying neural field estimated from observations.}

>>>>

The single layer homogeneous neural field model in the most general form is described by the equation
\begin{equation}
	\label{FullDoubleIntModel} \underbrace{v\left(\mathbf{r},t\right)}_{\text{field}} =
	\int_{-\infty}^t 
	\underbrace{h\left(t - t'\right)}_{\text{synapse}} \int_\Omega
	\parham{\theta}\underbrace{w\left(\mathbf{r},\mathbf{r}'\right)}_{\text{connectivity}} 
	\underbrace{f\left( v\left( \mathbf{r}',t' \right)\right)}_{\text{firing}}
	\, \mathrm{d}\mathbf{r}'\mathrm{d}t',
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ within the spatial domain $\Omega$ is denoted by $v\left(\mathbf r,t\right)$. The function $h(t)$ describes the post-synaptic response kernel, $w(\mathbf{r})$ describes the connectivity kernel and $f(v(\mathbf{r},t))$ describes the relationship between the mean membrane potential and the mean firing rate. The parameter $\theta$ can be thought of the synaptic efficiency or coupling strength, where $\theta>0$ and $\theta<0$ describes excitation and inhibition, respectively. There are various forms of $h(t)$ that have been proposed in the literature, with different levels of complexity. These are shown in Figure~\ref{fig:Modelcomponents}. Similarly, both the connectivity kernel and the activity function (firing rate) can take various forms. These are also shown in Figure~\ref{fig:Modelcomponents}. The form of $w(\mathbf{r})$ typically describes an Laplacian (exponential) or Gaussian decay in connectivity strength from any point in the field. By assuming the form of $h(t)$ is the same across layers (scales) the Amari style neural field model is formed, incorporating a multilayer structure into the connectivity kernel. In this way, the connectivity kernel typically takes the form of a Mexican hat (Gaussian shape within layer connectivity) or Wizard hat (Laplacian shape within layer) as depicted in Figure~\ref{fig:Modelcomponents}.

@@

The $m$-th  order cardinal B-spline function is defined by the recurrence relation \citep{Chui1992} 
\begin{equation}
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)\,\mathrm{d}r',
\label{SplineConvolutionIntegral}
\end{equation}
where $m>1$, $\ast$ denotes convolution, and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$, also known as indicator function, i.e.
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Following this, a B-spline of any order, $N_m(r)$, can be computed using the recursive expression defined by \citep{DeBoor2001}
\begin{equation}\label{eq:MRA-DoBoorFormula}
 N_{m}\left(r\right)=\frac{r}{m-1}N_{m-1}\left(r\right)+\frac{m-r}{m-1}N_{m-1}\left(r-1\right) \quad m>1.
 \end{equation}
The most commonly used form of splines is the cubic spline $\left(m=4\right)$, comprised of third degree polynomials added together at the joining points, with an explicit expression given by
\begin{align}
3!N_{4}\left(r\right)=
\begin{cases}
r^3 & \text{if $ 0\le r\le1$}, \\
4-12r+12r^2-3r^3 & \text{if $1\le r\le2$}, \\
-44+60r-24r^2+3r^3 & \text{if $2\le r\le3$}, \\
64-48r+12r^2-r^3 & \text{if $3\le r\le4$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{align}
One important feature of the B-spline functions is the \emph{partition of unity} property, meaning that the unity can be expressed as a linear sum of B-splines, i.e
\begin{equation}
	\sum_{l}N_m(r+l)\equiv1
	\end{equation}
	This is important as the underlying field can be reconstructed when its saturated without introducing numerical errors which is not the case for example in the case of Gaussian basis functions.

@@

By exploiting \eqref{eq:BsplineConvolution} and \eqref{eq:BsplineInnerProduct}, the integrals in \eqref{eq:Lambdax}, \eqref{eq:Lambdatheta} and \eqref{eq:CovMatrix} can be computed analytically. In constructing $\boldsymbol\Lambda_{x}$, it is important to note that B-spline scaling and wavelet functions possess the following orthogonality properties \citep{Unser1993}: 
\begin{equation}
  \left\langle \varphi_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\neq j_2
 \label{eq:MRA-PsiPsiOrthogonality} 
 \end{equation}
 \begin{equation}
  \left\langle N_{m;j_1,l_1}(r),\varphi_{m;j_2,l_2}(r)\right\rangle =0  \quad \mathrm{for} \quad j_1\leq j_2.
 \label{eq:MRA-PhiPsiOrthogonality}
 \end{equation}
  \parham{Note from \eqref{eq:MRA-PsiPsiOrthogonality} B-spline wavelets are semiorthogonal as they are not orthogonal with respect to translation on a given scale.}

@@

\subsection{Frequency Analysis}
To determine the level of approximation in the model, detailed in the previous section, that is the number of basis functions required to reconstruct the field, $v_t(r)$, the frequency response of the B-spline function needs to be computed by applying the Fourier transform to \eqref{eq:N1convolutions}. To proceed the Fourier transform of $N_1(r)$ is first calculated, 
\begin{align}\label{eq:MRA-N1Fouriertransform}
\mathcal F(N_1(r))&=\int_{-\infty}^{+\infty}N_1(r)\mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\int_{0}^{1} \mathrm{exp}(-2\pi i \nu r)dr \nonumber \\
&=\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}.
\end{align}
The convolution theorem states that the Fourier transform of the convolution of functions is equal to the product of their Fourier transforms. Therefore, taking Fourier transform of \eqref{eq:N1convolutions} and substituting \eqref{eq:MRA-N1Fouriertransform} for $\mathcal F(N_1(r)) $ yields
\begin{align}\label{eq:MRA-NmFouriertransform}
\mathcal F(N_m(r))=\left(\frac{1-\mathrm{exp}(-2\pi i \nu)}{2\pi i\nu}\right)^m.
\end{align}
\dean{Need to say how the FT is used somewhere here before proceeding.} \parham{The Fourier transform of the $m$-th order B-spline allows the calculation of the spatial frequency response of the wavelets using \eqref{eq:MRA-TwoScalepair2} from which the level of approximation, $j$, can be determined. This way we choose wavelets upto level $j$ which fully describe the spatial frequency contents of the underlying neural field estimated from observations.}
